Ludwig User Tutorial

### 2017-10-20 docs/tutorial/tutorial.tex
Minor edits in tutorial
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3170 40876441-3496-0410-a613-a2fca35b2ae2

This tutorial document takes new users of the \texttt{Ludwig} code through the process of how to obtain 
and build the code before outlining some short tutorial exercises where simulations 
of simple example cases will be run and visualised.
It is assumed that the reader is familar with UNIX-based operationg systems and has 
a basic knowledge of hydrodynamics, complex fluids, and to some extent statistical physics. 
This knowledge will be required to make sense of the input and output involved in using the code.

The system requirements are an up-to-date version of a C compiler as well as working installations 
of Apache's Subversion (SVN) version control system, the visualisation tool
ParaView (\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.

@@

\subsection{Configuration and Build}

@@

\begin{enumerate}
\item Go to the directory \texttt{/ludwig/trunk/}: \\
\begin{lstlisting}
$ cd ludwig/trunk
\end{lstlisting}
\item Copy the configuration file \texttt{lunix-gcc-default.mk} to the \texttt{/trunk} directory 
and rename to \texttt{config.mk}: \\
\begin{lstlisting}
$ cp config/lunix-gcc-default.mk ./config.mk
\end{lstlisting}
Note that you may need to specify the path to the MPI wrapper compiler in this file.
\item Issue \texttt{make} in \texttt{/targetDP}: \\
\begin{lstlisting}
$ cd targetDP/
$ make 
\end{lstlisting}
This creates dummy functions for calls to the targetDP library.
\item Issue \texttt{make} in \texttt{/mpi\_s}: \\
\begin{lstlisting}
$ cd ../mpi_s/
$ make 
\end{lstlisting}
This creates dummy functions for calls to the MPI-library,
which is required for serial compilation.
\item Edit \texttt{Makefile} in \texttt{/src}. To switch off 
the assertions set \texttt{OPTS = -DNP\_D3Q6 -DNDEBUG}. This is
important for production runs (see also comment in \texttt{Makefile}): \\
\begin{lstlisting}
$ cd ../src/
$ vim Makefile
...

@@

OPTS = -DNP_D3Q6 -DNDEBUG
LIBS = -L../targetDP -ltarget -lm
INC = -I. -I ../targetDP
...
\end{lstlisting}
\item To compile the code in serial issue: \\
\begin{lstlisting}
$ make serial
\end{lstlisting} 
To compile the parallel version of the code issue: \\
\begin{lstlisting}
$ make mpi
\end{lstlisting} 
This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}.
\end{enumerate}

@@

\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids. Firstly compile the code in serial and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}. These are the file required for the initialisation and visualisation of colloids respectively.
\item In the file, \texttt{colloid\_init.c} set the flags for:
\begin{enumerate}
\item \texttt{NMC = 0},
\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make colloid_init
$ ./colloid_init
\end{lstlisting}
this produces a file, \texttt{config.cds.init.001-001} which should be copied into your working directory.
\item In the file, \texttt{extract\_colloids.c} set:
\begin{enumerate}
\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
\item \texttt{cds\_with\_v  = 1} \\ which will output the velocities and positions of the colloids,
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make extract_colloids
\end{lstlisting}
and copy the executable \texttt{extract\_colloids} into your working directory along with the 
usual post-processing scripts, \texttt{vtk\_extract.c} and \texttt{vtk\_extract\_script.py} 
scripts. 
\item Next, run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item For the post-processing of this system, in \texttt{vtk\_extract.c}, set:
\begin{enumerate}
\item \texttt{input\_binary\_  = 1} \\ due to binary output from code,
\item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
\item \texttt{output\_lc\_op\_   = 1} \\ to visualise the liquid crystal order parameter,
\item \texttt{output\_lc\_dir\_  = 1} \\ to visualise the liquid crystal director field,
\item \texttt{output\_lc\_biop\_ = 0} \\ since the liquid crystals are not biaxial,
\end{enumerate}
\item and in \texttt{vtk\_extract\_script.py},
\begin{enumerate}
\item \texttt{vel = 1}, \\ to visualise the velocity field,
\item \texttt{q = 1}, \\ since there are liquid crystals present,
\item \texttt{colcds = 0} and 
\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
\end{enumerate}
\item After running the python script as before, \texttt{.csv} files with the colloid data should 
be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
\item The system can now be visualised, again using ParaView:
\begin{enumerate}
\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
\texttt{op-q-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:

>>>>

This tutorial document takes new users of the \texttt{Ludwig} code through the process of how to obtain 
and build the code before outlining some short tutorial exercises where simulations 
of simple example cases will be run and visualised.
It is assumed that the reader is familar with UNIX-based operating systems and has 
a basic knowledge of hydrodynamics, complex fluids, and to some extent statistical physics. 
This knowledge will be required to make sense of the input and output involved in using the code.

The system requirements are an up-to-date version of a C compiler as well as working installations 
of a text editor like \texttt{vi} or \texttt{emacs}, Apache's Subversion (SVN) version control system, the visualisation tool
ParaView (\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.

@@

\subsection{Configuration and Build}\label{configbuild}

@@

\begin{enumerate}
\item Go to the directory \texttt{/ludwig/trunk/}: \\
\begin{lstlisting}
$ cd ludwig/trunk
\end{lstlisting}
\item Copy the configuration file \texttt{lunix-gcc-default.mk} to the \texttt{/trunk} directory 
and rename to \texttt{config.mk} (both steps exectued below at once): \\
\begin{lstlisting}
$ cp config/lunix-gcc-default.mk ./config.mk
\end{lstlisting}
Note that you may need to specify the path to the MPI wrapper compiler in this file.
\item Issue \texttt{make} in \texttt{/targetDP}: \\
\begin{lstlisting}
$ cd targetDP/
$ make 
\end{lstlisting}
This creates dummy functions for calls to the targetDP library.
\item Issue \texttt{make} in \texttt{/mpi\_s}: \\
\begin{lstlisting}
$ cd ../mpi_s/
$ make 
\end{lstlisting}
This creates dummy functions for calls to the MPI-library,
which is required for serial compilation.
\item Edit \texttt{Makefile} in \texttt{/src}. To switch off 
the assertions set \texttt{OPTS = -DNP\_D3Q6 -DNDEBUG}. This is
important for production runs (see also comment in \texttt{Makefile}): \\
\begin{lstlisting}
$ cd ../src/
$ vim Makefile
...

@@

OPTS = -DNP_D3Q6 -DNDEBUG
LIBS = -L../targetDP -ltarget -lm
INC = -I. -I ../targetDP
...
\end{lstlisting}
\item To compile the code in serial issue: \\
\begin{lstlisting}
$ make serial
\end{lstlisting} 
To compile the parallel version of the code issue: \\
\begin{lstlisting}
$ make mpi
\end{lstlisting} 
This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}. 
Before going from a serial to a parallel compilation or vice versa
you should clean the directory by issuing
\begin{lstlisting}
$ make clean
\end{lstlisting}

\end{enumerate}

@@

\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids. Firstly compile the code in serial (see section \ref{configbuild})and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}. These are the file required for the initialisation and visualisation of colloids respectively.
\item In the file, \texttt{colloid\_init.c} set the flags for:
\begin{enumerate}
\item \texttt{NMC = 0},
\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make colloid_init
$ ./colloid_init
\end{lstlisting}
this produces a file, \texttt{config.cds.init.001-001} which should be copied into your working directory.
\item In the file, \texttt{extract\_colloids.c} set:
\begin{enumerate}
\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
\item \texttt{cds\_with\_v  = 1} \\ which will output the velocities and positions of the colloids,
\end{enumerate}
\item Now issue: \\
\begin{lstlisting}
$ make extract_colloids
\end{lstlisting}
and copy the executable \texttt{extract\_colloids} into your working directory along with the 
usual post-processing scripts, \texttt{vtk\_extract.c} and \texttt{vtk\_extract\_script.py} 
scripts. 
\item Next, run the code in your working directory by issuing the command:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe input
\end{lstlisting}
\item For the post-processing of this system, in \texttt{vtk\_extract.c}, set:
\begin{enumerate}
\item \texttt{input\_binary\_  = 1} \\ due to binary output from code,
\item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
\item \texttt{output\_lc\_op\_   = 1} \\ to visualise the liquid crystal order parameter,
\item \texttt{output\_lc\_dir\_  = 1} \\ to visualise the liquid crystal director field,
\item \texttt{output\_lc\_biop\_ = 0} \\ since the liquid crystals are not biaxial,
\end{enumerate}
\item and in \texttt{vtk\_extract\_script.py},
\begin{enumerate}
\item \texttt{vel = 1}, \\ to visualise the velocity field,
\item \texttt{q = 1}, \\ since there are liquid crystals present,
\item \texttt{colcds = 0} and 
\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
\end{enumerate}
\item After running the python script as before, \texttt{.csv} files with the colloid data should 
be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
\item The system can now be visualised, again using ParaView:
\begin{enumerate}
\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
\texttt{op-q-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:
